Selective Sequential Model Selection

### 2015-12-05 paper.tex
lots of small changes

\JTcomment{Maybe we should write $k_F$ instead of $k_0$? Or $k_0(Y,F)$?}\\
\WFcomment{$\hat k_F$ is already the ForwardStop estimator. But $k_0(Y,F)$ is good.}\\

@@

\subsubsection{Forward Stepwise Paths and Greedy Likelihood Ratio Tests}
Let $\ell(\theta; Y)$ denote the log-likelihood for model
$M_\infty$. A generic {\em forward stepwise} algorithm proceeds as follows: we begin with some fixed set $E_0$ (such as an intercept-only model), then at step $k=1,\ldots,d$, we set
\begin{equation}
j_k = \argmax_j \; \sup \left\{\ell(\theta; Y):\; \theta\in\Theta(E_{k-1} \cup \{j\})\right\}, \quad \text{ and } \quad
E_k = E_{k-1} \cup \{j_k\}.
\end{equation}
That is, at each step we select the next variable to maximize the likelihood of the next fitted model.

@@

For simplicity, we have made two assumptions: that only one variable is added at each step, and that the set of candidate variables we choose from is the same in each step. It is relatively straightforward to relax either assumption, but we do not pursue such generalizations here.

@@

\paragraph{Remark.} In its usual formulation, the lasso coefficients for linear regression minimize a penalized RSS criterion. If we replace $-\ell$ with any strictly decreasing function of the log-likelihood, all of the results in this article hold without modification.

@@

Similarly, the error control guarantees of \citet{gsell2013sequential} and \citet{li2015accumulation} do not directly apply to case where the null hypotheses are random. However, we recover these guarantees if we have {\em conditional} independence and uniformity of null $p$-values given $k_0$:
\begin{equation}\label{eq:indepCond_random_k0}
  \P(p_{k+1} \leq \alpha_{k+1}, \ldots, p_d \leq \alpha_d
  \mid p_1, \ldots, p_k, k_0 = k ) \eqAS \left(\prod_{i=k+1}^d \alpha_i\right) .
\end{equation}
For the sake of brevity, we will say that $p$-value sequences satisfying~\eqref{eq:indepCond_random_k0} are {\em independent on nulls}.

@@

\begin{proof}
  Noting that $\{k_0=k\}$ is $\F_k$-measurable, it is enough to show that 
  \begin{equation}\label{eq:jointSuper2}
  \P_F\left(p_{k+1}\leq \alpha_{k+1}, \ldots, p_d \leq \alpha_d, \;
    \mid\; \sF_k\right)\1_{\{k_0\leq k\}} \leqAS \left(\prod_{i=k+1}^d
  \alpha_i\right)\1_{\{k_0\leq k\}}
  \end{equation}
  We now prove~\eqref{eq:jointSuper2} by induction. Define $B_i = \{p_i \leq \alpha_i\}$. The base case is
  \[
  \P_F\left(B_d \mid \sF_{d-1}\right)1_{\{k_0 \leq d-1\}} \leqAS \alpha_d
1_{\{k_0 \leq d-1\}},
  \]
  which is true by construction of $p_d$. 
  For the inductive case, note that 
  \begin{align*}
    \P_F\left(B_{k+1}, \ldots, B_d
      \mid \sF_k\right)1_{\{k_0 \leq k\}} 
    &\eqAS \E_F\bigg[ 1_{B_{k+1}} 
    \P_F\left(B_{k+2}, \ldots, B_d
      \mid \sF_{k+1}\right)1_{\{k_0 \leq k+1\}}
    \mid \sF_k\bigg]1_{\{k_0 \leq k\}}\\
    &\leqAS \P_F\left[ B_{k+1}
      \mid \sF_k\right]1_{\{k_0 \leq k\}}\prod_{i=k+2}^d \alpha_i \\
    &\leqAS \left(\prod_{i=k+1}^d \alpha_i \right)1_{\{k_0 \leq k\}} .
  \end{align*}
Lastly, if the $p_k$ are exact then the inequalities above become equalities, implying uniformity and joint independence.
\end{proof}

@@

We begin by proving both paths satisfy the SSP
\begin{proposition}\label{prop:forwardSSP}
Forward stepwise paths and ever-active regularization paths both satisfy the subpath sufficiency principle.
\end{proposition}

@@

Saturated-model tests perform inference on selected linear functionals of $\mu$. In the context of sequential linear model selection,  we can use the saturated-model framework to test the incremental null 
\[
H^{\text{inc}}:\; \theta_{j_k}^{E_k} = 0,
\]
where $\theta^E$ minimizes \smash{$\|\mu - X_E\theta\|_2$}.

Let $\eta_k\in \R^n$ denote the vector for which \smash{$\theta_{j_k}^{E_k} = \eta_k'\mu$}, and let \smash{$\proj_{\eta_k}^\perp$} denote the projection operator into the space orthogonal to $\eta_k$. The UMPU saturated-model test for $\eta_k'\mu$ is based on the distribution
\begin{equation*}
\L\left( \eta_k'Y \mid E_{k-1}, \;j_k, \;\proj_{\eta_k}^\perp Y \right).
\end{equation*}

While the test statistic $X_{j_k}'Y$ is measurable with respect to $\sF_k$, the null distribution we compare it against is not: it depends on \smash{$\proj_{\eta_k}^\perp Y$}, which is not measurable with respect to $\sF_k$. As a result, saturated-model $p$-values constructed as in~\eqref{eq:satModel} are in general {\em not} independent on nulls.

@@

\subsubsection{Testing Against the Next Model}\label{sec:identify}
One option for selected-model inference is to perform a selective test of $M_{k-1}$ against $M_{k}\setminus M_{k-1}$, conditioning on both models. Conditioning on $(M_{k-1},M_k)$ is equivalent to conditioning on $(E_{k-1}, j_k)$. If $\sigma^2$ is known, the test is based on
\begin{equation}\label{eq:selModel_cond}
\L\left(X_{j_k}'Y \mid E_{k-1}, \; j_k, \; X_{E_{k-1}}'Y\right).
\end{equation}
If $\sigma^2$ were unknown we would also need to condition on $\|Y\|^2$. We can construct $p$-values from the UMPU, equal-tailed, or one-sided test based on~\eqref{eq:selModel_cond}; the only difference is in how much of the overall level $\alpha$ is apportioned to the right or left tail \citep{fithian2014optimal}. If we use the construction in~\eqref{eq:selModel_cond} after a subpath-sufficient path algorithm, the resulting $p$-values are always guaranteed to have independence on nulls. This is because the test statistic $X_{j_k}'Y$ is measurable with respect to $\sF_k$, and so is the null distribution we compare it against, which depends on $E_{k-1}, j_k,$ and \smash{$U_{k-1} = X_{E_{k-1}}'Y$}.

Note the contrast between~\eqref{eq:selModel_cond} and~\eqref{eq:satModel}. In~\eqref{eq:selModel_cond}, we only need to condition on an $|E_{k-1}|$-dimensional projection of $Y$, whereas in~\eqref{eq:satModel} we condition on an $(n-1)$-dimensional projection of $Y$. Conditioning on more information can sap the power of a test; as we will see in Sections~\ref{sec:bivariate} and~\ref{sec:sparseReg}, the saturated-model tests tend to pay a heavy price for this extra conditioning.

@@

%In the case of known $\sigma^2$, the greedy likelihood ratio statistic is the max-$z$ statistic: the maximum of $z_{j,E_{k-1}}$, the largest multivariate $z$-statistic of any held-out variable, controlling for the variables in the current model.

For the most part we do not pursue tests like those in Section~\ref{sec:identify}, which condition on the next model $M_k$. It is not necessary to condition on $M_k$ to obtain the FDR and FWER guarantees that we want to establish, and so we follow the rule of thumb that we should condition on as little information as possible. In Section~\ref{sec:sparseReg}, we empirically compare the max-$t$ test with the test that uses the same test statistic but also conditions on the identity of the next variable $j_k$. We call the resulting test the {\em max-$t$-identify test}. Its performance in our simulation is similar to that of the max-$t$ test.

@@

Selected-model inference is typically more computationally involved than saturated-model inference. Because the saturated-model test conditions on $n-1$ dimensions, the resulting distribution in~\eqref{eq:satModel} is nothing more than a truncated univariate Gaussian random variable, and inference can be carried out analytically. By contrast, selected-model tests typically require sampling from a truncated multivariate Gaussian distribution of dimension $p-|E_{k-1}|$.

For some applications, a major conceptual advantage of the saturated-model approach is that, when $M(E)$ is misspecified, $\theta^{E}$ is still a well-defined quantity on which we can perform exact inference. For example, a 95\% confidence interval for $\theta_{j_k}^{E_k}$ could cover that coefficient with probability 95\% even if $M(E_k)$ is misspecified. In the setting of sequential model selection, this advantage is less important: we are not constructing intervals for $\theta_{j_k}^{E_k}$; rather, the goal of the test at step $k$ is to reject $M_{k-1}$ and move on if it is demonstrably incompatible with the data.

There are several major statistical benefits to selected-model inference. First, selected-model inference allows us to drop the assumption that $\sigma^2$ is known. This is not an option for the saturated model because conditioning on both $\|Y\|^2$ and \smash{$\proj_{\eta_k}^\perp Y$} results in a degenerate conditional distribution for $\eta_k'Y$. Second, we have seen in Section~\ref{sec:pValsIndep} that tests of the form~\eqref{eq:selModel_cond} or~\eqref{eq:selModel_marg} yield independent $p$-values under general conditions, allowing us to apply  the sequential stopping rules of~\citet{gsell2013sequential} and \citet{li2015accumulation}. Finally, and perhaps most importantly, selected-model tests can be dramatically more powerful than saturated-model tests, as we illustrate now with an extended example.

\begin{example}[Bivariate Regression with Identity Design]\label{ex:bivariate}
  Consider forward stepwise selection in a regression model with $n=p=2$, with known $\sigma^2=1$ and identity design 
\[
X = I_2=\begin{pmatrix} 1 & 0 \\ 0 & 1\end{pmatrix}.
\] 
The forward stepwise path and the lasso path both select $j_1=1$ if and only if $|Y_1|>|Y_2|$. The selection event $A_1=\{|Y_1| > |Y_2|\}$ is shown in yellow in Figure~\ref{fig:bv_condSets}. If $j_1=1$ the first two models are
\[
M_0:\; Y\sim \cN(0,I_2), \quad \text{ and } \quad
M_1:\; Y \sim \cN\left(\binom{\mu_1}{0}, \; I_2\right).
\]
The selected-model test at step 1 compares $Y_1$ to its distribution under $M_0$ conditional on $A_1$, a test of $H_0:\;\mu_1=0$ in model $M_1$.\footnote{In this very simple example, the max-$z$, max-$z$-identify, and next-entry tests all coincide.} By contrast, the saturated-model test is a test of $H_0:\; \mu_1=0$ in the model $M_{\text{sat}}:\; Y \sim \cN(\mu, I_2)$. Thus, it must condition on $Y_2$ to eliminate the nuisance parameter $\mu_2$, comparing $Y_1$ to its null distribution given $A_1$ and the observed value of $Y_2$.

Figure~\ref{fig:bv_condSets} shows the conditioning sets for each model when $Y=(2.9, 2.5)$. Beside it, Figure~\ref{fig:bv_nullDists} shows the null conditional distribution of the test statistic $Y_1$ for each test. The $p$-values for the selected and saturated models are 0.007 and 0.3, respectively. These two plots are reproduced from \citet{fithian2014optimal}, in which the same example was presented in less detail.
\end{example}

@@

Results in \citet{fithian2014optimal} show that the selected-model test is strictly more powerful when the selected model is correct, i.e., when $\mu_2=0$. Figure~\ref{fig:bv_powCurves} shows the power curve for each test when $\mu_2=0$ (left panel) and $\mu_2=4$ (right panel). While the selected-model test is more powerful when the selected model is correct, the difference between the two is relatively small. The difference is much more pronounced when $\mu_2=4$. Note that if $\mu=(0,4)$, the incremental null is true, but the complete null is false. That is, the model $M_0 = M(\emptyset)$ is missing an important signal variable, but the missing variable is not $X_1 = (1,0)$. Because the saturated-model test is a valid test of the incremental null, its power is $\alpha=0.05$. By contrast, the selected-model test rejects about half of the time when $\mu=(0,4)$, detecting the fact that $M_0$ is a poor fit to the data.

@@

Saturated-model $p$-values are generically non-independent. Continuing with Example~\ref{ex:bivariate}, Table~\ref{tab:bv_twoWayTable} shows a two-way contingency table for the saturated-model $p$-values $(p_1(Y), p_2(Y))$, binned into cells of height and width 0.2, simulated under the global null $\mu=0$. Because $k_0=0$, both $p$-values are uniform by construction, but the saturated-model $p$-values are strongly dependent, with correlation $-0.48$. By contrast, the selected-model $p$-values $(p_1,p_2)$ are independent under the global null, a consequence of Theorem~\ref{thm:suffCond}.

@@

Until this point, we have deferred discussing the question of how to actually compute the $p$-values for the max-$t$ test, next-entry test, or any of the other tests we have discussed. In each case, the $p$-value at step $k$ comes from a test that rejects when some test statistic $T_k$ is large compared to the law
\begin{equation}\label{eq:genDist}
\L(T_k(Y) \,\mid\, M_{k-1}(Y), U_{k-1}(Y)),
\end{equation}
under the null hypothesis that $F\in M_{k-1}$. Because $U_{k-1}$ is sufficient for $Y$ under the null, this law is completely known. Thus, we have reduced computation of the $p$-value at step $k$ to a sampling problem: if we can sample values of $Y$ from its null distribution given $M_{k-1}(Y), U_{k-1}(Y)$, we can numerically approximate the $p$-value to whatever precision we desire.

In many cases it is easy to sample $Y$ given $U_{k-1}$, in which case computational difficulties only arise insofar as the selection event $\{M_{k-1}(Y)=M\}$ is difficult to condition on. Due to the generality in which we have posed the problem, we cannot give a general prescription for how to carry out this sampling. However, we give some details for the important case of linear regression in Appendix~\ref{sec:linRegComputation}.

@@

Remarkably, however, in both forward-stepwise and lasso linear regression, the computational problem actually becomes easier when we move to the sequential setting considered here. Indeed, Section~\ref{sec:constraints} shows that for both forward-stepwise regression and lasso linear regression --- and their respective generalizations to exponential family models --- conditioning on the event $\{E_{k-1}=E\}$, for any $k$, only introduces $2p$ linear constraints, taking the form of a rectangular constraint on the sufficient statistics for the inactive variables.

\subsection{Constraints}\label{sec:constraints}
We now show that the conditioning polytope can be expressed with only $2(p-k)$ constraints in forward stepwise regression and the lasso. This increases the speed of both accept/reject and hit-and-run sampling for our problem.

We begin by proving the result for forward stepwise linear regression, followed by a more general result for the forward stepwise likelihood path in exponential families.

\begin{theorem}
  Assume the model path $M_{0:d}$ is obtained by forward stepwise 
  linear regression. Then, for a candidate active set $E$ of size $k$, 
  the set $A = \{E_k = E, \;X_E'Y = u\}$ is characterized 
  exactly by the constraints $X_E'Y=u$ and
  \[
  v_j^-(E,u) < X_j'Y < v_j^+(E,u), \quad\forall j \notin E,
  \]
  for $v_j^-$ and $v_j^+$ given explicitly in (\ref{eq:vMinus_FS}--\ref{eq:vPlus_FS}).  Thus, $A$ corresponds exactly to a set with $2(p-k)$ linear inequality constraints and $k$ linear equality constraints on $Y$.
\end{theorem}

\RJTcomment{Can't we get rid of the strict inequalities above? The proof writes things in terms of nonstrict inequalities, and I think it makes sense to drop them; in the case of ties, $E$ is still a valid selected model.}
\WFcomment{Thanks for noticing the inconsistency. }

@@

\begin{theorem}
  Assume that $M_\infty$ is an exponential family model
  of the form
  \[
  Y \sim \exp\{ \theta'U(y) - \psi(\theta) \}\,d\nu(y),
  \]
  with $\Theta \sub \R^p$ convex, and assume that the model path $M_{0:d}$ is given by the forward stepwise likelihood path with
\begin{equation}
j_k = \argmax_j \;\sup \left\{\ell(\theta; Y):\; \theta\in\Theta(E_{k-1} \cup \{j\})\right\}, \quad \text{ and } \quad
E_k = E_{k-1} \cup \{j_k\}.
\end{equation}
  Then, for a candidate active set $E$ of size $k$, the set $A = \{E_k = E, \;U_E = u\}$ is characterized exactly by the constraints $U_E = u$ and
  \[
  v_j^-(E,u) \leq U_j(Y) \leq v_j^+(E,u), \quad\forall j \notin E,
  \]
  for $v_j^-$ and $v_j^+$ given in (\ref{eq:vMinus_FSL}--\ref{eq:vPlus_FSL}). Thus, $A$ corresponds exactly to a set with $2(p-k)$ linear inequality constraints and $k$ linear equality constraints on $U(Y)$.
\end{theorem}

\begin{proof}
  For a set $E\sub \{1,\ldots,p\}$ define
  \[
  \ell_E^*(U(Y)) 
  = \sup\left\{\theta'U(Y) - \psi(\theta) :\; \theta\in \Theta(E) \right\}
  = \sup\left\{\theta_E'U_E(Y) - \psi_E(\theta_E):\; \theta\in\Theta(E) \right\},
  \]
  where $\psi_E:\; \R^{|E|} \to \R$ is defined such that $\psi(\theta)=\psi_E(\theta_E)$ if $\theta_{\{1,\ldots p\}\setminus E}=0$. Note that $\ell_E^*$ is the convex dual of the convex function $\psi_E$, so $\ell_E^*$ is also convex. \JTcomment{It would be convex even if $\psi_E$ were not convex. No? I also think $U$ in the definition of $\ell^*_E$ can be just $U(Y)$.} \RJTcomment{I would call this the ``convex conjugate'' or just the ``conjugate'' of $\ell$; and yes this will always be convex.  It doesn't require knowing anything about $\psi_E$ or even $\Theta$.} At stage $i<k$, the next variable to enter is the one that maximizes \smash{$\ell_{E_i\cup \{j\}}^*(U(Y))$}.

>>>>

@@

\subsubsection{Forward Stepwise Paths and Greedy Likelihood Ratio Tests}
Let $\ell(\theta; Y)$ denote the log-likelihood for model
$M_\infty$. A generic {\em forward stepwise} algorithm proceeds as follows: we begin with some fixed set $E_0$ (such as an intercept-only model), then at step $k=1,\ldots,d$, we set
\begin{equation}
j_k = \argmax_j \; \sup \left\{\ell(\theta; Y):\; \theta\in\Theta(E_{k-1} \cup \{j\})\right\}, \quad \text{ and } \quad
E_k = E_{k-1} \cup \{j_k\}.
\end{equation}
That is, at each step we select the next variable to maximize the likelihood of the next fitted model (in the case of ties, we could either choose randomly or select both variables).

@@

For simplicity, we have implicitly made two assumptions: that only one variable is added at each step, and that the set of candidate variables we choose from is the same in each step. It is relatively straightforward to relax either assumption, but we do not pursue such generalizations here.

@@

\paragraph{Remark.} In its usual formulation, the lasso coefficients for linear regression minimize a penalized RSS criterion. If we replace $-\ell$ with any strictly decreasing function of the log-likelihood such as RSS, all of the results in this article hold without modification.

@@

Similarly, the error control guarantees of \citet{gsell2013sequential} and \citet{li2015accumulation} do not directly apply to case where the null hypotheses are random. However, we recover these guarantees if we have {\em conditional} independence and uniformity of null $p$-values given $k_0$:
\begin{equation}\label{eq:indepCond_random_k0}
  \P_F(p_{k+1} \leq \alpha_{k+1}, \ldots, p_d \leq \alpha_d
  \mid p_1, \ldots, p_k) \eqAS \left(\prod_{i=k+1}^d \alpha_i\right) \quad \text{on } \{k_0(Y, F) = k\}.
\end{equation}
For the sake of brevity, we will say that $p$-value sequences satisfying~\eqref{eq:indepCond_random_k0} are {\em independent on nulls}.

@@

\begin{proof}
  Noting that $\{k_0=k\}$ is $\F_k$-measurable, it is enough to show that 
  \begin{equation}\label{eq:jointSuper2}
  \P_F\left(p_{k+1}\leq \alpha_{k+1}, \ldots, p_d \leq \alpha_d, \;
    \mid\; \sF_k\right)\1_{\{k_0\leq k\}} \leqAS \left(\prod_{i=k+1}^d
  \alpha_i\right)\1_{\{k_0\leq k\}}
  \end{equation}
  We now prove~\eqref{eq:jointSuper2} by induction. Define $B_i = \{p_i \leq \alpha_i\}$. The base case is
  \[
  \P_F\left(B_d \mid \sF_{d-1}\right)1_{\{k_0 \leq d-1\}} \leqAS \alpha_d
1_{\{k_0 \leq d-1\}},
  \]
  which is true by construction of $p_d$. 
  For the inductive case, note that 
  \begin{align*}
    \P_F\left(B_{k+1}, \ldots, B_d
      \mid \sF_k\right)1_{\{k_0 \leq k\}} 
    &\eqAS \E_F\bigg[ 1_{B_{k+1}} 
    \P_F\left(B_{k+2}, \ldots, B_d
      \mid \sF_{k+1}\right)1_{\{k_0 \leq k+1\}}
    \mid \sF_k\bigg]1_{\{k_0 \leq k\}}\\
    &\leqAS \P_F\left[ B_{k+1}
      \mid \sF_k\right]1_{\{k_0 \leq k\}}\prod_{i=k+2}^d \alpha_i \\
    &\leqAS \left(\prod_{i=k+1}^d \alpha_i \right)1_{\{k_0 \leq k\}} .
  \end{align*}
Lastly, if the $p_k$ are exact then the inequalities above become equalities, implying uniformity and mutual independence.
\end{proof}

@@

We begin by proving both paths satisfy the SSP:
\begin{proposition}\label{prop:forwardSSP}
Forward stepwise paths and ever-active regularization paths both satisfy the subpath sufficiency principle.
\end{proposition}

@@

Saturated-model tests perform inference on selected linear functionals of $\mu$. In the context of sequential linear model selection,  we can use the saturated-model framework to test the incremental null \smash{$H^{\text{inc}}:\; \theta_{j_k}^{E_k} = 0$}, where $\theta^E$ minimizes \smash{$\|\mu - X_E\theta\|_2$}.

Let $\eta_k\in \R^n$ denote the vector for which \smash{$\theta_{j_k}^{E_k} = \eta_k'\mu$}, and let \smash{$\proj_{\eta_k}^\perp$} denote the projection operator into the space orthogonal to $\eta_k$. The UMPU saturated-model test for $\eta_k'\mu$ is based on the distribution \smash{$\L\left( \eta_k'Y \mid E_{k-1}, \;j_k, \;\proj_{\eta_k}^\perp Y \right)$}, or equivalently
\begin{equation}\label{eq:satModel}
\L\left( X_k'Y \mid E_{k-1}, \;j_k, \;\proj_{\eta_k}^\perp Y \right)
\end{equation}

While the test statistic $X_{j_k}'Y$ is measurable with respect to $\sF_k$, the null distribution we compare it against is not: it depends on \smash{$\proj_{\eta_k}^\perp Y$}, which is not measurable with respect to $\sF_k$. As a result, saturated-model $p$-values are in general {\em not} independent on nulls.

@@

\subsubsection{Testing Against the Next Model}\label{sec:identify}
One option for selected-model inference is to perform a selective test of $M_{k-1}$ against $M_{k}\setminus M_{k-1}$, conditioning on both models. Conditioning on $(M_{k-1},M_k)$ is equivalent to conditioning on $(E_{k-1}, j_k)$. If $\sigma^2$ is known, the test is based on
\begin{equation}\label{eq:selModel_cond}
\L\left(X_{j_k}'Y \mid E_{k-1}, \; j_k, \; X_{E_{k-1}}'Y\right).
\end{equation}
If $\sigma^2$ were unknown we would also need to condition on $\|Y\|^2$. We can construct $p$-values from the UMPU, equal-tailed, or one-sided test based on the one-parameter exponential family~\eqref{eq:selModel_cond}; the only difference is in how much of the overall level $\alpha$ is apportioned to the right or left tail \citep{fithian2014optimal}. If we use the construction in~\eqref{eq:selModel_cond} after a subpath-sufficient path algorithm, the resulting $p$-values are always guaranteed to have independence on nulls. This is because the test statistic $X_{j_k}'Y$ is measurable with respect to $\sF_k$, as is the reference null distribution.

Note the contrast between~\eqref{eq:selModel_cond} and~\eqref{eq:satModel}. In~\eqref{eq:selModel_cond}, we only need to condition on an $|E_{k-1}|$-dimensional projection of $Y$, whereas in~\eqref{eq:satModel} we condition on an $(n-1)$-dimensional projection of $Y$. Conditioning on more information can sap the power of a test; as we will see in Sections~\ref{sec:bivariate} and~\ref{sec:sparseReg}, the saturated-model tests often pay a heavy price for this extra conditioning.

@@

We will not pursue tests like those in Section~\ref{sec:identify}. It is not necessary to condition on $M_k$ to obtain the FDR and FWER guarantees that we want to establish, and so we follow the general principle of conditioning on as little information as possible. In Section~\ref{sec:sparseReg}, we empirically compare the max-$z$ test with the test that uses the same test statistic but also conditions on the identity of the next variable $j_k$. In our simulation this test, which we call the {\em max-$z$-identify test}, performs similarly to the max-$z$ test.

@@

Selected-model inference is typically more computationally involved than saturated-model inference. Because the saturated-model test conditions on $n-1$ dimensions, the resulting distribution in~\eqref{eq:satModel} is nothing more than a truncated univariate Gaussian random variable, and inference can be carried out analytically. By contrast, selected-model tests typically require MCMC sampling from a truncated multivariate Gaussian distribution of dimension $p-|E_{k-1}|$.

For some applications, a major conceptual advantage of the saturated-model approach is that, even when $M(E)$ is misspecified, $\theta^{E}$ is still a well-defined quantity on which we can perform exact inference. For example, a 95\% confidence interval for $\theta_{j_k}^{E_k}$ could cover that coefficient with probability 95\% even if $M(E_k)$ is misspecified. In the setting of sequential model selection, this advantage is less important: at an intermediate step $k$, we are not constructing intervals for $\theta_{j_k}^{E_k}$ but rather deciding whether to reject $M_{k-1}$ and move on. However, the saturated-model framework could be quite helpful after selection is finished, if we want to construct intervals for the least-squares parameters for the selected active set $E_{\hk}$.

There are several major statistical benefits to selected-model inference. First, selected-model inference allows us to drop the assumption that $\sigma^2$ is known. This is not possible for the saturated model because conditioning on both $\|Y\|^2$ and \smash{$\proj_{\eta_k}^\perp Y$} results in a degenerate conditional distribution for $\eta_k'Y$. Second, we have seen in Section~\ref{sec:pValsIndep} that tests of the form~\eqref{eq:selModel_cond} or~\eqref{eq:selModel_marg} yield independent $p$-values under general conditions, allowing us to apply  the sequential stopping rules of~\citet{gsell2013sequential} and \citet{li2015accumulation}. Finally, and most importantly, selected-model tests can be dramatically more powerful than saturated-model tests, as we illustrate in the next example.

\begin{example}[Bivariate Regression with Identity Design]\label{ex:bivariate}
  Consider forward stepwise selection in a regression model with $n=p=2$, with known $\sigma^2=1$ and identity design 
\[
X = I_2=\begin{pmatrix} 1 & 0 \\ 0 & 1\end{pmatrix}.
\] 
The forward stepwise path and the lasso path both select $j_1=1$ if and only if $|Y_1|>|Y_2|$. The selection event $A_1=\{|Y_1| > |Y_2|\}$ is shown in yellow in Figure~\ref{fig:bv_condSets}. On $A_1$, the first two models are
\[
M_0:\; Y\sim \cN(0,I_2), \quad \text{ and } \quad
M_1:\; Y \sim \cN\left(\binom{\mu_1}{0}, \; I_2\right).
\]
The selected-model test at step 1 compares $Y_1$ to its distribution under $M_0$, the global null, conditional on $A_1$.\footnote{In this very simple example, the max-$z$, max-$z$-identify, and next-entry tests are all equivalent.} By contrast, the saturated-model test is a test of $H_0:\; \mu_1=0$ in the model $M_{\text{sat}}:\; Y \sim \cN(\mu, I_2)$. Thus, it must condition on $Y_2$ to eliminate the nuisance parameter $\mu_2$, comparing $Y_1$ to its null distribution given $A_1$ and the observed value of $Y_2$.

Figure~\ref{fig:bv_condSets} shows the conditioning sets for each model when $Y=(2.9, 2.5)$. Beside it, Figure~\ref{fig:bv_nullDists} shows the null conditional distribution of the test statistic $Y_1$ for each test. The $p$-values for the selected and saturated models are 0.007 and 0.3, respectively. Figure~\ref{fig:bv} is reproduced from \citet{fithian2014optimal}, where the same example was presented in less detail.
\end{example}

@@

Results in \citet{fithian2014optimal} show that the selected-model test is strictly more powerful when the selected model is correct, i.e., when $\mu_2=0$. Figure~\ref{fig:bv_powCurves} shows the power curve for each test when $\mu_2=0$ (left panel) and $\mu_2=4$ (right panel). While the selected-model test is more powerful when the selected model is correct, the difference between the two is relatively small. The difference is much more pronounced when $\mu_2=4$. Note that if $\mu=(0,4)$, the incremental null is true, but the complete null is false. That is, the global null model $M_0$ is missing an important signal variable, but the missing variable is not $X_1 = (1,0)$. Because the saturated-model test is a valid test of the incremental null, its power is $\alpha=0.05$. By contrast, the selected-model test rejects about half of the time when $\mu=(0,4)$, successfully detecting the poor fit of $M_0$ to the data.

@@

Saturated-model $p$-values are generically non-independent. Continuing with Example~\ref{ex:bivariate}, Table~\ref{tab:bv_twoWayTable} shows a two-way contingency table for the saturated-model $p$-values $(p_1(Y), p_2(Y))$, binned into cells of height and width 0.2, simulated under the global null $\mu=0$. Because $k_0=0$, both $p$-values are uniform by construction, but the saturated-model $p$-values are strongly dependent, with correlation $-0.48$. By contrast, the selected-model $p$-values $(p_1,p_2)$ are independent under the global null.

@@

Until this point, we have deferred discussing how to compute the reference distribution for the max-$t$ test, next-entry test, or any of the other tests we have discussed. In each case, the $p$-value at step $k$ comes from a test that rejects when some test statistic $T_k$ is large compared to the law
\begin{equation}\label{eq:genDist}
\L(T_k(Y) \,\mid\, M_{k-1}(Y), U_{k-1}(Y)),
\end{equation}
under the null hypothesis that $F\in M_{k-1}$. Because $U_{k-1}$ is sufficient for $Y$ under the null, this law is completely known. Thus, we have reduced computation of the $p$-value at step $k$ to a sampling problem: if we can sample values of $Y$ from its null distribution given $M_{k-1}(Y), U_{k-1}(Y)$, we can numerically approximate the $p$-value to whatever precision we desire.

In many cases it is easy to sample $Y$ given $U_{k-1}$, in which case computational difficulties only arise insofar as the selection event $\{M_{k-1}(Y)=M\}$ is difficult to condition on, or the test statistic $T_k(Y)$ is difficult to evaluate. Due to the generality in which we have posed the problem, we cannot give a general prescription for how to carry out this sampling. However, we give some details for the important case of linear regression in Appendix~\ref{sec:linRegComputation}.

\subsection{Constraints}\label{sec:constraints}

@@

Remarkably, however, in both forward-stepwise and lasso linear regression, the computational problem actually becomes easier when we move to the sequential setting considered here. We now show that for both forward-stepwise regression and lasso linear regression --- and their respective generalizations to exponential family models --- conditioning on the event $\{E_{k-1}=E, U_k=u\}$, for any $k, E,$ and $u$, only introduces $2(p-k)$ linear inequality constraints, amounting to a rectangular constraint for the sufficient statistics of excluded variables. This increases the speed of both accept/reject and hit-and-run sampling for our problem. We begin by proving the result for forward stepwise linear regression, followed by a more general result for the forward stepwise likelihood path in exponential families. To avoid trivialities, we assume throughout that there are almost surely no ties. This is true, for example, in linear regression if $X$ is in general position.

\begin{theorem}
  Assume the model path $M_{0:d}$ is obtained by forward stepwise 
  linear regression. Then, for a candidate active set $E$ of size $k$, 
  the set $A = \{E_k = E, \;X_E'Y = u\}$ is characterized 
  exactly by the constraints $X_E'Y=u$ and
  \[
  v_j^-(E,u) \leq X_j'Y \leq v_j^+(E,u), \quad\forall j \notin E,
  \]
  for $v_j^-$ and $v_j^+$ given explicitly in (\ref{eq:vMinus_FS}--\ref{eq:vPlus_FS}).  Thus, $A$ corresponds exactly to a set with $2(p-k)$ linear inequality constraints and $k$ linear equality constraints on $Y$.
\end{theorem}

@@

\begin{theorem}
  Assume that $M_\infty$ is an exponential family model
  of the form
  \[
  Y \sim \exp\{ \theta'U(y) - \psi(\theta) \}\,d\nu(y),
  \]
  with $\Theta \sub \R^p$ convex, and assume that $M_{0:d}$ is the forward stepwise likelihood path with
\begin{equation}
j_k = \argmax_j \;\sup \left\{\ell(\theta; Y):\; \theta\in\Theta(E_{k-1} \cup \{j\})\right\}, \quad \text{ and } \quad
E_k = E_{k-1} \cup \{j_k\}.
\end{equation}
  Then, for a candidate active set $E$ of size $k$, the set $A = \{E_k = E, \;U_E = u\}$ is characterized exactly by the constraints $U_E = u$ and
  \[
  v_j^-(E,u) \leq U_j(Y) \leq v_j^+(E,u), \quad\forall j \notin E,
  \]
  for $v_j^-$ and $v_j^+$ given in (\ref{eq:vMinus_FSL}--\ref{eq:vPlus_FSL}). Thus, $A$ corresponds exactly to a set with $2(p-k)$ linear inequality constraints and $k$ linear equality constraints on $U(Y)$.
\end{theorem}

\begin{proof}
  For a set $E\sub \{1,\ldots,p\}$ define
  \[
  \ell_E^*(U(Y)) 
  = \sup\left\{\theta'U(Y) - \psi(\theta) :\; \theta\in \Theta(E) \right\}
  = \sup\left\{\theta_E'U_E(Y) - \psi_E(\theta_E):\; \theta\in\Theta(E) \right\},
  \]
  where $\psi_E:\; \R^{|E|} \to \R$ is defined such that $\psi(\theta)=\psi_E(\theta_E)$ for $\theta\in \Theta(E)$. Note that $\ell_E^*$ is the convex conjugate of the function $\psi_E$. For step $i<k$, write
  \[
  C_i^* = \max_{j \notin E_i} \;\ell_{E_i\cup \{j\}}^*(U(Y)).
  \]
  On $A$, $C_i^*$ is attained at $j=j_{i+1}\in E$, the next variable to enter. For $j\notin E$, the condition for $j$ {\em not} to enter at step $i+1$ is \smash{$\ell_{E_i\cup \{j\}}^*(U(Y)) \leq C_i^*$}. Due to the SSP, on the set $\{E_k=E\}$ all of the quantities $j_{i+1}$, $E_i$, and $C_i^*$ are functions of $U_E(Y)$.

  Because $\ell_{E_i\cup \{j\}^*}(U)$ depends only on $U_{E_i}$ and $U_j$, the former of which is fixed at $u_{E_i}$, we can consider it as a one-dimensional convex function of $U_j$; hence any sublevel set is an interval with possibly infinite endpoints. Thus
  \begin{equation}\label{eq:noEnterBounds_FSL}
  \{U_E=u, \; \ell_{E_i\cup \{j\}}^*(U(Y)) \leq C_i^*\} 
  = \{ U_E=u, \; v_{j,i}^-(E,u) \leq U_j \leq v_{j,i}^+(E,u)\},
  \end{equation}
  where $v_{j,i}^{-}$ and $v_{j,i}^+$ are the two values of $U_j$ such that \smash{$\ell_{E_i\cup \{j\}}^*(U)=C_i^*$}, or else $\pm\infty$.
